**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-06 submission/manuscript.Rmd
intro done for now

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States (cite). CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen. These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite). The proximal and distal human colon are comprised of different microhabitats that likely enrich for different microbial membership at each site. Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite). Further, our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite).

These studies have been limited though by the sample collection methods. Nearly all human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite). While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite). Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but has been proven to disrupt the normal gut microbiome

- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
- specific keystone species are associated with CRC tumors
- Central Question: do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations

- What data is needed to answer this question? first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa

this is a really good paper, good for framing intro and discussion : Numerical ecology validates a biogeographical distribution and gender-based effect on mucosa-associated bacteria along the human colon

- what methods are used to get the data? unprepped colonoscopy

- What analysis must be applied? - diversity analysis and machine learning to pull out small variations/differences among the noisy data set.

- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool

>>>>

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States (cite). CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen. These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite). The proximal and distal human colon are comprised of different microhabitats that likely enrich for different microbial membership at each site. Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.

Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (cite). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (cite). These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (cite Niel).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.

Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite). While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite). Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (Jalanka 2015, Harrell 2012). As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

In this study we sought to combine these approaches to characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon. Our design uses unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers. To get around the noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.
